Spatial population structure determines extinction risk in climate-induced range shifts

### 2019-05-01 Manuscript/AmNat_ReSubmission.tex
Updated the model scripts and manuscript for the AmNat revisions

\textit{Manuscript elements}: Figure~1, figure~2, figure~3, online appendices~A and B (including figures~A1-A3,  tables~A1-A4, and figures~B1-B11). Figure~1 and figure~3 are to print in color.

@@

\section*{Abstract}
Climate change is an escalating threat facing populations around the globe, necessitating a robust understanding of the ecological and evolutionary mechanisms dictating population responses. However, populations do not respond to climate change in isolation, but rather in the context of their existing ranges. In particular, spatial population structure within a range (e.g. trait clines, starkness of range edges, etc.) likely interacts with other ecological and evolutionary processes during climate-induced range shifts. Here, we use an individual-based model to explore the interacting roles of these factors in range shift dynamics. We show that increased spatial population structure (driven primarily by a steeper slope in the trait optimum gradient) severely increased a population's extinction risk. Further, and contrary to expectations, we show that evolution of heightened dispersal during range shifts was unable to rescue faltering populations. Rather, a population's fate during climate change was determined by the composition of dispersal phenotypes defining the population at equilibrium (i.e. before the onset of rapid climate change); only populations consisting of highly dispersive individuals survived. Our results demonstrate that dispersal evolution alone may be insufficient to save a range shifting population and that spatial population structure can substantially increase extinction risk in range shifts.

@@

Large-scale population movements have been studied for decades in the context of range expansions, leading to a robust understanding of both the ecological~\citep{hastings2005spatial} and evolutionary~\citep{shine2011evolutionary, excoffier2009genetic} mechanisms shaping such expansions. For example, while range expansion speed can be well approximated by a combination of the species' intrinsic growth rate and dispersal ability~\citep{fisher1937wave, hastings2005spatial}, recent research demonstrates that evolution in these traits can increase both the mean and variance of expansion speed through time~\citep{szHucs2017rapid, shaw2015dispersal, phillips2015evolutionary}. In particular, laboratory experiments have shown spatial genetic structure to play an important role in the dynamics of range expansions, even over short time periods~\citep{weiss2017rapid,ochocki2017rapid}. While some range expansions (e.g. of invasive or reintroduced species) start from small founding populations lacking any initial spatial structure, others are characterized by far more complex initial spatial population structure (e.g. humans expanding out of Africa or tree expansion following glacial retreat). Climate-driven range shifts fall in the latter category, as populations respond to climate change in the context of preexisting spatial population structure. For example, species' ranges may be characterized by spatial clines in abundance and genetic patterns. Such clines can form in response to a number of extrinsic or intrinsic factors including adaptation to environmental gradients~\citep{kirkpatrick1997evolution}, altered demographic rates throughout the range~\citep{holt2005theoretical}, or interactions with other species~\citep{case2000interspecific}. Importantly for climate-driven range shifts, the factors influencing these spatial patterns will likely shift with the changing climate, causing simultaneous range contraction and expansion at opposite range margins. Thus, climate-driven range shifts will be affected both by the eco-evolutionary dynamics known to impact range expansions and the moving constraints on spatial population structure impacted by changing climatic conditions.

@@

Here, we assess the interaction of two mechanisms responsible for spatial population structure (edge type and an environmental gradient in a trait optimum) with trait evolution in sexually reproducing populations undergoing range shifts. We develop an individual-based model capable of producing a wide variety of spatial population structures in which males and females are defined by two genetically determined traits, thus allowing for both evolutionary and ecological responses to climate change. One trait determines dispersal ability while the second defines an individual's environmental niche. Using this model, we vary both the environmental gradient in the niche optimum and the starkness of the range edge to ascertain how they interact with each other and with the process of trait evolution to impact a population's ability to track a changing climate. Previous research suggests evolution of heightened dispersal during the range shift might alleviate the extinction risk of moving populations~\citep{hargreaves2015fitness, boeye2013more, henry2013eco}, but we predict steep environmental gradients and stark range edges may inhibit such evolution and thus increase extinction risk.

\section*{Methods}
A full description of the individual-based model using the Overview, Design concepts, and Details protocol~\citep{grimm2010odd} is available in Appendix A, while we present a brief summary here. Population dynamics occurred within discrete habitat patches embedded in a two dimensional lattice in which environmental conditions varied along the $x$ dimension but remained constant along the $y$ dimension (Fig. A1). Landscapes were unbounded in the $x$ dimension but defined by a fixed width and wrapping boundaries in the $y$ dimension. The optimum environmental niche value changed linearly along the $x$ dimension, thus allowing for the intrinsic formation of stable range boundaries when the optimum changed rapidly enough~\citep{kirkpatrick1997evolution, alleaume2006geographical, polechova2015limits, polechova2018sky}. However, to examine the dynamics of ranges in which the niche optimum did not change rapidly enough to form stable range limits, we additionally imposed extrinsic range limits to prevent continuous adaptation and spread of the population~\citep{alleaume2006geographical, garcia1997genetic}. Specifically, we systematically altered the decline in patch carrying capacities from the range core to the edge~\citep{alleaume2006geographical, henry2013eco, bocedi2014rangeshifter, mustin2009dynamics} in such a way that we could directly manipulate the starkness of the decline. Previous research has shown that using a decline in intrinsic growth rate as opposed to carrying capacity may impact extinction risk but does not alter the patterns of dispersal evolution during climate change~\citep{henry2013eco}. To maintain generality, we do not assume a specific mechanism behind the decline in carrying capacity, but it could represent a variety of range limiting mechanisms such as physiological limits to adaptation, the effects of competition, or underlying resource distributions~\citep{sexton2009evolution, holt2005theoretical, case2000interspecific}. Thus, the $x$ dimension defined the environmental context of the population and the $y$ dimension allowed for variation in population dynamics under identical environmental conditions. To simulate climate change, the patch carrying capacities and the niche optimum gradient shifted at a constant rate along the $x$ dimension. Generations were non-overlapping and consisted of discrete dispersal and reproduction phases (Fig. A2).

Individuals were characterized by two traits (dispersal and an environmental niche), each defined by a set of $5$ quantitative diploid loci. While the number of loci was arbitrary, $5$ was chosen as a compromise between computational restrictions and the likely polygenic nature of such complex traits. The dispersal trait defined an individual's expected dispersal distance, assuming an exponential dispersal kernel. An individual's realized dispersal distance was then drawn from the dispersal kernel and dispersal direction was random and unbiased. Dispersal occurred in continuous space from the center of an individual's current patch and the individual's new patch was then determined by the mapping from continuous space to discrete patches (see Appendix A). An individual's environmental niche value determined its fitness according to the local niche optimum. The closer an individual's niche value to the local optimum, the higher the individual's realized fitness. Reproduction within each patch occurred via a stochastic implementation of the classic Ricker model~\citep{ricker1954stock, melbourne2008extinction}, scaled by the mean fitness of the patch. Parental pairs formed via random sampling of the local population (with replacement) weighted by individual fitness such that individuals with a close match of their niche value to the local optimum produced more offspring on average. Thus, the model used a mixture of  hard selection (realized population growth declined with maladaptation relative to the niche optimum) and soft selection (probability of producing offspring depended on fitness relative to other individuals) for the evolutionary dynamics~\citep{wallace1975hard}. Allele inheritance was subject to mutation and assumed no linkages among loci. The mutation process was designed such that mutational input per generation was independent of the number of loci (see Appendix A) and with parameters corresponding to previous estimates from the literature~\citep{gilbert2017local}.

We varied parameter values to explore the interacting roles of the slope of the niche optimum gradient and range edge starkness (Table A2) in forming spatial population structure at equilibrium and driving the subsequent eco-evolutionary dynamics of range shifts. Specifically, we considered a fully factorial combination of: (1) a flat, shallow, and steep niche optimum gradient, (2) shallow, moderate, and stark declines in carrying capacity at the range edge, and (3) slow, moderate, and fast speeds of climate change. This yielded a total of $27$ different scenarios, each explored with $200$ simulations. Each simulation ran for $2150$ generations with stable climate conditions for the first $2000$ to reach a spatial equilibrium, followed by $100$ generations of climate change and a final $50$ generations of stable conditions. Figure 1 shows an example of a single population responding to a moderate speed of climate change. For each scenario, we evaluated the role of the equilibrium spatial population structure and dispersal evolution in the dynamics of the range shifting populations. We primarily discuss simulations assuming a moderate speed of climate change in the main text, but present the results for slow and fast speeds of climate change in Appendix B.

We calculated dispersal evolution in each patch throughout the landscape as the change in mean dispersal phenotype from the beginning of the period of climate change to the end. For this analysis, we defined individual patches by their relative location within the range rather than with their fixed spatial coordinates (e.g. leading edge vs. core populations). Due to local extinctions, not all patches were occupied at the end of the period of climate change. To quantify dispersal evolution in these patches, we used data from the last generation in which the population had at least $10$ individuals. Changes in mean dispersal phenotype were calculated by subtracting the initial mean dispersal phenotype from the value at the end of climate change (or at the last generation of at least $10$ individuals occupying the patch in the case of population extinctions); positive values indicate an increase in the mean dispersal phenotype. All simulations and data processing were performed in R version $3.4.4$~\citep{team2000r} and the code is available at (links are available from the journal office).

\section*{Results}
In all scenarios, some populations shifted their ranges in response to climate change. However, the proportion of populations that failed to track the changing climate (and subsequently went extinct) depended on the spatial characteristics of the range. Populations defined by a steep niche optimum gradient and by stark declines in carrying capacity at the range edge experienced the greatest probability of extinction due to climate change (quantified by the proportion of simulated populations to go extinct through time; Fig. 2). While both aspects of a population's range influenced extinction probabilities, the niche optimum gradient drove more dramatic changes to extinction risk, with steeper gradients causing severe increases in the probability of extinction during climate change. We varied both parameters widely, doubling the slope of the niche optimum gradient from the shallow to steep scenario and increasing the parameter defining range edge starkness by a factor of $100$ from shallow to stark edges (Table A2). This suggests that the niche optimum gradient may be the stronger driver of extinction risk during climate-induced range shifts across a wide region of parameter space and corresponding biological scenarios. Additionally, as expected, the pace of climate change also influenced extinction probabilities with faster climate change corresponding to greater extinction risk (Fig. B1 \& B2). However, this effect was independent of the niche optimum gradient and range edge starkness in determining extinction probability during range shifts.

In accordance with previous results~\citep{garcia1997genetic}, populations at the range edge had lower fitness than central populations at equilibrium and this effect was amplified with more gradual range edges. Counterintuitively, populations that survived climate change tended to be characterized by even greater reductions in fitness at the range edges at equilibrium compared to populations that went extinct (Fig. B3-B5). While discernible in all simulations with a non-zero slope in the niche optimum gradient, this pattern was most evident in the scenarios defined by a gradual range edge. As expected, there was no spatial variation in fitness for populations with no variation in the niche optimum across space. Despite the spatial variation in fitness in some scenarios, variance in relative fitness within a patch was relatively low (about $0.3$ across all scenarios). This implies that (1) the reduction in fitness at the edge caused by the niche optimum gradient was relatively uniform across all individuals, and (2) as a result, there was relatively low variance in reproductive success in these populations, meaning that evolution at the edge was not driven by only a handful of higher fitness individuals.

Dispersal evolution is predicted to play a key role in aiding populations as they shift to track a changing climate. While some individual simulations confirmed these expectations with average dispersal phenotypes increasing through time (e.g. Fig. 1), examining all simulations from each experimental scenario revealed no differences in the magnitude or direction of dispersal evolution between successful and extinct populations (Fig. 3a\&b). Populations in all parameter combinations experienced both increases and decreases in average dispersal phenotypes, with all distributions of observed changes in dispersal phenotypes centered on $0$ (Fig. B3-B5). Further, calculating the coefficient of variation (CV) in dispersal genotypes over the $y$ dimension, in which environmental conditions did not vary, revealed that edge populations had between $3$ and $4$ times higher CVs than core populations across all scenarios, indicating genetic diversity at the edge was not a limiting factor in dispersal evolution. The similarity in evolved changes in dispersal between surviving and extinct populations suggests that dispersal evolution alone cannot explain which populations successfully tracked moving conditions and which became extinct.

Instead, the distribution of dispersal phenotypes at equilibrium played a key role in determining a population's fate. A range of dispersal phenotypes evolved in populations over the $2000$ generations of stable climatic conditions in response to the niche optimum gradient and range edge starkness (Fig. B9-B11). Populations that survived climate change were composed primarily of individuals with heightened dispersal phenotypes (Fig. 3c\&d). Previous research has demonstrated that defining the range edges via a decline in the intrinsic growth rate (as opposed to carrying capacity as done here) resulted in less dispersive phenotypes at the range edge~\citep{henry2013eco}, meaning extinction risks would be even higher under such a scenario. Comparing the full distribution of equilibrium dispersal phenotypes present in a given experimental scenario to the distribution of phenotypes just from surviving populations revealed a threshold value delineating individuals from surviving versus extinct populations. Comparison of the different experimental scenarios revealed this threshold to be constant for a given speed of climate change (Fig. B9-B11). To explain this phenomenon, we used the well-known approximation for the speed of an expanding population, $2\sqrt{rD}$~\citep{fisher1937wave, hastings2005spatial}, in which $r$ is the intrinsic growth rate and $D$ is the diffusion coefficient, to calculate the dispersal phenotype necessary to produce an expansion wave exactly matching the speed of climate change in our simulations (see the model description in Appendix A). This estimated dispersal phenotype matched the observed threshold value distinguishing surviving from extinct populations in all experimental scenarios (Figures B9-B11, vertical dashed line). Thus, surviving populations in each scenario happened to be the lucky few already composed primarily of individuals with dispersal phenotypes capable of spreading at the pace of climate change, rather than populations in which heightened dispersal evolved over time in response to climate change. However, this threshold effect weakened slightly in scenarios with a slow speed of climate change, indicating a potential role for dispersal evolution if the climate were to change at a slow enough rate (Figure B11).

\section*{Discussion}
Range shifts due to climate change represent a global threat to biodiversity and much recent research has focused on exploring the underlying ecological and evolutionary dynamics of such range shifts to inform conservation efforts. We developed an individual-based model to explore the eco-evolutionary dynamics of climate-induced range shifts in sexually reproducing, diploid populations with both dispersal and environmental niche traits defined by multiple loci. In contrast, previous models have focused on a subset of these factors: ecological dynamics (e.g.~\citep{brooker2007modelling}), evolution in a single trait only (e.g.~\citep{atkins2010local, henry2013eco}), and relatively simple genetic scenarios (e.g. single-locus haploid genetics in asexual populations~\citep{boeye2013more, hargreaves2015fitness}).  Here, we tested the generality of previous results predicting an important role for dispersal evolution in range shifts~\citep{boeye2013more, henry2013eco} and the interplay of eco-evolutionary dynamics under increased levels of biological complexity. Specifically, we demonstrated the role of spatial population structure, driven by a niche optimum gradient and range edge starkness, in determining extinction risk for range shifting populations via impacts on the equilibrium distribution of dispersal phenotypes and environmental niche values.

Our results suggest that populations characterized by local adaptation to a spatially varying trait optimum and by stark range edges will be less able to track changing climatic conditions (Fig. 2). A survey of the scientific literature found evidence for local adaption in approximately $71\%$ of studies, suggesting a high prevalence of local adaptation in natural populations~\citep{hereford2009quantitative}. While it is difficult to exactly map adaptation to the niche optimum gradient to empirical measures of local adaptation, the parameters defining the steepest gradient used here resulted in the intrinsic formation of stable range boundaries, as seen in previous theoretical studies~\citep{kirkpatrick1997evolution, alleaume2006geographical, polechova2015limits, polechova2018sky}, suggesting they provide reasonable approximations of empirical patterns. Further, a recent meta-analysis of $1400$ bird, mammal, fish, and tree species found no evidence for consistent declines in abundance towards range edges~\citep{dallas2017species}, suggesting many species exhibit similar abundances at the edge and center of their ranges similar to the starkest range edges imposed in our study. While some of these patterns could represent a publication bias, for example against negative results in studies of local adaptation, combined with our results they suggest many species will face elevated extinction risks in climate-induced range shifts due to their spatial population structure.

Our results emphasize the importance of the equilibrium distribution of dispersal phenotypes (i.e. prior to the onset of rapid climate change) in determining a population's extinction risk during climate change (Fig. 3c\&d). Populations primarily composed of high dispersal phenotypes at equilibrium successfully tracked changing climate conditions, while populations of lower dispersal phenotypes lagged behind the changing conditions to eventually go extinct. Importantly, the threshold dividing the equilibrium dispersal phenotypes of successful and extinct populations was constant across all scenarios for a given speed of climate change (Fig. B9-11). Thus, the difference in survival probability among scenarios was driven by the effects of the niche optimum gradient and range edge starkness on the evolution of dispersal ability throughout the range. Scenarios with a steep niche optimum gradient selected for lower dispersal phenotypes at equilibrium due to the potential mismatch of dispersing individuals' niche phenotype and their new location~\citep{kirkpatrick1997evolution}. Similarly, a stark range edge selected for lower dispersal due to the risk of dispersing into unsuitable habitat~\citep{shaw2014population, shaw2019evolution}. Previous research has documented a similar reduction in dispersal phenotypes due to an explicit mortality cost of dispersal~\citep{kubisch2013predicting}, whereas the costs to dispersal in our model result from the niche optimum gradient and range edge starkness. Such a cost to dispersal results in lower dispersal phenotypes at equilibrium, hampering a population's ability to successfully track a changing climate. Importantly, dispersal evolution during climate change was unable to counter the influence of equilibrium spatial population structure on extinction dynamics.

>>>>

\textit{Manuscript elements}: Figures~1-6, online appendices~A and B (including figures~A1-A3,  tables~A1-A4, and figures~B1-B10). All figures should print in color.

@@

\section*{Abstract}
Climate change is an escalating threat facing populations around the globe, necessitating a robust understanding of the ecological and evolutionary mechanisms dictating population responses. However, populations do not respond to climate change in isolation, but rather in the context of their existing ranges. In particular, spatial population structure within a range (e.g. trait clines, starkness of range edges, etc.) likely interacts with other ecological and evolutionary processes during climate-induced range shifts. Here, we use an individual-based model to explore the interacting roles of several such factors in range shift dynamics. We show that increased spatial population structure (driven primarily by a steeper environmental gradient) severely increases a population's extinction risk. Further, we show that while evolution of heightened dispersal during range shifts can aid populations in tracking changing conditions, it can also interact negatively with adaptation to the environmental gradient, leading to reduced fitness and contributing to the increased extinction risk observed in populations structured along steep environmental gradients. Our results demonstrate that the effect of dispersal evolution on range shifting populations is dependent on environmental context and that spatial population structure can substantially increase extinction risk in range shifts.

@@

Large-scale population movements have been studied for decades in the broader context of range expansions, leading to a robust understanding of both the ecological~\citep{hastings2005spatial} and evolutionary~\citep{shine2011evolutionary, excoffier2009genetic} mechanisms shaping such expansions. For example, while range expansion speed can be well approximated by a combination of the species' intrinsic growth rate and dispersal ability~\citep{fisher1937wave, hastings2005spatial}, recent research demonstrates that evolution in these traits, such as spatial sorting leading to increased dispersal at expanding edges~\citep{shine2011evolutionary}, can increase both the mean and variance of expansion speed through time~\citep{szHucs2017rapid, shaw2015dispersal, phillips2015evolutionary}. In particular, laboratory experiments have shown spatial genetic structure to play an important role in the dynamics of range expansions, even over short time periods~\citep{weiss2017rapid, ochocki2017rapid}. While some range expansions (e.g. of invasive or reintroduced species) start from small founding populations lacking any initial spatial structure, others are characterized by far more complex initial spatial population structure (e.g. humans expanding out of Africa or expansions following glacial retreat). Climate-driven range shifts fall in the latter category, as populations respond to climate change in the context of preexisting spatial population structure. For example, species' ranges may be characterized by spatial clines in abundance and genetic patterns. Such clines can form in response to a number of extrinsic or intrinsic factors including adaptation to environmental gradients~\citep{kirkpatrick1997evolution}, altered demographic rates throughout the range~\citep{holt2005theoretical}, or interactions with other species~\citep{case2000interspecific}. Importantly for climate-driven range shifts, the factors influencing these spatial patterns will likely shift with the changing climate, causing simultaneous range contraction and expansion at opposite range margins. Thus, climate-driven range shifts will be affected both by the eco-evolutionary dynamics known to impact range expansions and the moving constraints on spatial population structure impacted by changing climatic conditions.

@@

Here, we assess the interaction of two mechanisms responsible for spatial population structure (edge type and an environmental gradient in a trait optimum) with trait evolution in sexually reproducing populations undergoing range shifts. We developed an individual-based model capable of producing a wide variety of spatial population structures in which females and males were defined by two genetically determined traits, thus allowing for both evolutionary and ecological responses to climate change. One trait determined dispersal ability while the second allowed for adaptation to the environmental gradient. Using this model, we varied both the environmental gradient in the trait optimum and the starkness of the range edge to ascertain how they interact with each other and with the process of trait evolution to impact a population's ability to track a changing climate. Previous research suggests evolution of heightened dispersal during the range shift might alleviate the extinction risk of moving populations~\citep{hargreaves2015fitness, boeye2013more, henry2013eco}, but we predicted steep environmental gradients and stark range edges would inhibit such evolution and thus increase extinction risk.

\section*{Methods}
A full description of the individual-based model using the Overview, Design concepts, and Details protocol~\citep{grimm2010odd} is available in Appendix A, while we present a brief summary here. Population dynamics occurred within discrete habitat patches embedded in a two dimensional lattice in which environmental conditions varied along the $x$ dimension but remained constant along the $y$ dimension (Fig. A1). Landscapes were unbounded in the $x$ dimension but defined by a fixed width and wrapping boundaries in the $y$ dimension. Abiotic conditions varied along the $x$ dimension of landscapes in two ways. First, a linear environmental gradient determined the optimum phenotype in a quantitative trait, driving adaptation throughout the range. Second, the patch carrying capacities systematically declined from the range center to both edges along the $x$ dimension. The starkness of this decline could be altered to create different types of range edges (Fig. A3). While an environmental gradient in a quantitative trait optimum is sufficient to produce stable range limits under certain conditions~\citep{kirkpatrick1997evolution, alleaume2006geographical, polechova2015limits, polechova2018sky}, we included the decline in carrying capacities to enforce stable range limits when the environmental gradient was too shallow to produce them~\citep{alleaume2006geographical, garcia1997genetic}. A wide variety of other range limiting mechanisms have been proposed, including interspecific interactions~\citep{case2000interspecific, price2009evolutionarily}, Allee effects~\citep{keitt2001allee}, and spatial variation in demographic parameters~\citep{holt2005theoretical}. Given empirical evidence for many different range limiting mechanisms~\citep{gaston2009geographic}, the inclusion of both an environmental gradient in the trait optimum and the decline in carrying capacities allowed the model to capture a wide variety of spatial population structures, including ranges not structured by adaptation to an environmental gradient, with minimal additional assumptions. Previous research has shown that using a decline in intrinsic growth rate as opposed to carrying capacity may impact extinction risk but does not alter the patterns of dispersal evolution during climate change~\citep{henry2013eco}. To maintain generality, we do not assume a specific mechanism behind the decline in carrying capacity, but it could represent a variety of range limiting mechanisms such as physiological limits to adaptation, the effects of competition, or underlying resource distributions~\citep{sexton2009evolution, holt2005theoretical, case2000interspecific, price2009evolutionarily}. Thus, the $x$ dimension defined the environmental context of the population and the $y$ dimension allowed for variation in population dynamics under identical environmental conditions. To simulate climate change, the patch carrying capacities and the environmental gradient shifted at a constant rate along the $x$ dimension. Generations were non-overlapping and consisted of discrete dispersal and reproduction phases (Fig. A2).

Individuals were characterized by two traits, each defined by a set of $5$ quantitative diploid loci. While the number of loci was arbitrary, $5$ was chosen as a compromise between computational restrictions and the likely polygenic nature of such complex traits. The first trait defined an individual's expected dispersal distance, assuming an exponential dispersal kernel. An individual's realized dispersal distance was then drawn from the dispersal kernel and dispersal direction was random and unbiased. Dispersal occurred in continuous space from the center of an individual's current patch and the individual's new patch was then determined by the mapping from continuous space to discrete patches (see Appendix A). The second trait allowed for adaptation to the local conditions along the environmental gradient. The environmental gradient determined the local phenotypic optimum in this trait and deviation from this optimum decreased individual fitness. As this trait determined the environmental conditions in which individuals would experience maximal fitness, we refer to it as the niche trait when differentiating between it and the dispersal trait. Reproduction within each patch occurred via a stochastic implementation of the classic Ricker model~\citep{ricker1954stock, melbourne2008extinction}, scaled by the mean fitness of the patch. Parental pairs formed via random sampling of the local population (with replacement) weighted by individual fitness such that individuals with a close match of their niche trait to the local optimum produced more offspring on average. Thus, the model used a mixture of  hard selection (expected population growth declined with maladaptation to the environmental gradient) and soft selection (probability of producing offspring depended on fitness relative to other individuals) for the evolutionary dynamics~\citep{wallace1975hard}. Allele inheritance was subject to mutation and assumed no linkages among loci. The mutation process was designed such that mutational input per generation was independent of the number of loci (see Appendix A) and with parameters corresponding to previous estimates from the literature~\citep{gilbert2017local}.

We varied parameter values to explore the interacting roles of the slope of the environmental gradient and different types of range edges (Table A2) in forming spatial population structure at equilibrium and driving the subsequent eco-evolutionary dynamics of range shifts. Specifically, we considered a fully factorial combination of: (1) a flat, shallow, or steep environmental gradient, (2) gradual, moderate, or stark declines in carrying capacity at the range edge, and (3) slow, moderate, or fast speeds of climate change. This yielded a total of $27$ different scenarios, each explored with $200$ simulations, during which we tracked the mean phenotypes of both traits and their genetic variances in each patch through time. The first $2000$ generations of each simulation allowed populations to reach a spatial equilibrium, which we confirmed by assessing the spatial distributions of phenotypes and genetic variances in both traits for stability through time. On average, simulations in every scenario arrived at stable distributions in all values by generation $1500$. Following the initial $2000$ generations, we imposed $100$ generations of climate change by shifting the environmental gradient and patch carrying capacities at a constant rate. Figure 1 shows an example of a single population responding to a moderate speed of climate change. For each scenario, we evaluated the impact of the environmental gradient and edge type on trait evolution and extinction dynamics during the range shift. We primarily discuss simulations assuming a moderate speed of climate change in the main text, but present the results for slow and fast speeds of climate change in Appendix B. To assess spatial patterns in trait evolution, we focused primarily on variation along the $x$ dimension of simulated landscapes for two reasons: (1) to capture variation relevant to the changing environmental conditions and (2) because variation in this dimension was much greater than variation within the $y$ dimension in all scenarios. All simulations and data processing were performed in R version $3.4.4$~\citep{team2000r} and the code is available at (links are available from the journal office).

\section*{Results}
\subsection*{Extinction}
Extinction dynamics during the climate-driven range shifts varied widely among scenarios, ranging from no extinctions in some scenarios to the extinction of every simulated population in others. We quantified extinction probability in each scenario as the proportion of simulated populations to decline to extinction through time during the $100$ generations of climate change. Comparing extinction probabilities among scenarios revealed four prominent patterns. First, the slope of the environmental gradient strongly impacted extinction risk (Fig. 2). Scenarios with a flat gradient only resulted in extinctions at a fast speed of climate change coupled with a stark range edge (Fig. B2), while scenarios with a steep gradient experienced high extinction probabilities at even a slow speed of climate change (Fig. B1). Second, edge type also had an important effect on extinction probability, with stark edges increasing extinction risk over gradual and moderate edges. Interestingly, when scenarios with moderate and gradual range edges diverged in extinction probabilities, moderate edges consistently yielded lower extinction risk than scenarios with gradual edges (Fig. 2, B1, \& B2). Third, while both the environmental gradient and edge type influenced extinction probabilities, the environmental gradient drove more dramatic changes to extinction risk. We varied both parameters widely among scenarios, doubling the slope of the environmental gradient and increasing edge starkness by a factor of $100$ (Table A2). Thus, our results suggest the environmental gradient is the stronger driver of extinction risk during climate-induced range shifts across a wide region of parameter space and corresponding biological scenarios. Finally, as expected, the speed of climate change also influenced extinction probability with faster climate change corresponding to greater extinction risk (Fig. B1 \& B2).

\subsection*{Dispersal evolution} 
The distribution of dispersal phenotypes at spatial equilibrium varied across the $x$ dimension of simulated landscapes, with more dispersive phenotypes characterizing the range edges. While this pattern held generally across all scenarios, the difference between edge and core phenotypes was reduced in the presence of an environmental gradient and starker range edges, leading to lower dispersal phenotypes at the range margins in these populations (Fig. 3). At equilibrium, the distributions of dispersal phenotypes were symmetric, but spatial sorting during climate change not only increased dispersal phenotypes in surviving populations, but led to asymmetric distributions with the highest phenotypes at the leading edge. Despite these initial changes, the spatial distributions remained remarkably stable between $50$ and $100$ generations, regardless of the speed of climate change (Fig. 3, B3, \& B4). This stands in contrast to models of dispersal evolution in unbounded range expansions, which predict dispersal to continually increase as the expansion progresses~\citep{shaw2015dispersal, shine2011evolutionary, fronhofer2015eco}. In our model, the shifting environmental gradient and distribution of carrying capacities simultaneously exposed new habitat for colonization, thus allowing spatial sorting, and also reduced the fitness of individuals that dispersed too far, resulting in an upper limit on dispersal evolution.

Both the severity of the environmental gradient and edge starkness also determined the spatial distribution of genetic variance in the dispersal trait. Increasing the slope of the environmental gradient reduced the genetic variance in dispersal across the range, but did not alter the qualitative patterns. The type of range edge, on the other hand, qualitatively altered the spatial distribution of genetic variance in dispersal (Fig. 4). Scenarios with gradual and moderate edges displayed similar patterns of heightened variance at the range center compared to the margins, but stark edges resulted in the opposite pattern. This discrepancy makes sense as scenarios with stark declines in carrying capacity lacked the low abundance, and hence low genetic variance, populations that characterized the edges of scenarios with gradual or moderate declines in carrying capacity (Fig. A3). In all scenarios, genetic variance in dispersal decreased throughout the range in response to climate change (Fig. 4), consistent with selection for increased dispersal via spatial sorting. This pattern was qualitatively similar for all speeds of climate change (Fig. B5 \& B6).

\subsection*{Adaptation to the environmental gradient} 
Equilibrium values of the niche trait closely matched the environmental gradient in the range core and diverged at the edges (Fig. 5), likely due to a combination of gene swamping from abundant central populations~\citep{kirkpatrick1997evolution} and the loss of genetic diversity due to drift in small, peripheral populations~\citep{polechova2018sky, polechova2015limits}. In agreement with previous results, the starkness of the range edge influenced this pattern, with increased maladaptation of peripheral populations in scenarios with gradual edges~\citep{garcia1997genetic}. Across all scenarios, the clines in the niche trait became much shallower during the range shift (Fig. 5). As a result, the match between the environmental gradient and local trait values substantially deteriorated during climate change, resulting in reduced fitness across most of the range. This reduction in fitness, observed at all speeds of climate change (Fig. B7 \& B8), could partially explain the increased extinction probability observed for scenarios with a steep environmental gradient (Fig. 2).

In contrast to the dispersal trait, genetic variance in the niche trait at equilibrium actually increased with the slope of the environmental gradient (Fig. 6). This is likely due to the increased range of environmental conditions encountered by populations arranged along a steeper environmental gradient. The spatial distributions of genetic variance in the niche trait were qualitatively similar for scenarios with moderate and stark edges, peaking at the center of the range and declining towards the edges. However, scenarios with gradual edges displayed different patterns in the genetic variance of the niche trait, with a peak in variance occurring between the range center and each edge (Fig. 6). This pattern was lost during range shifts, though, leading to qualitatively similar distributions of genetic variance in the niche traits among different edge types (Fig. 6). In contrast to the loss of genetic variance observed in the dispersal trait during climate change, the magnitude of genetic variance in the niche trait during climate change was roughly equivalent to that characterizing equilibrium populations in most scenarios, regardless of the speed of climate change (Fig. B9 \& B10). Only in scenarios with gradual edges did genetic variance decrease slightly in response to climate change.

\section*{Discussion}
Range shifts due to climate change represent a global threat to biodiversity and recent research has focused on exploring the underlying ecological and evolutionary dynamics of such range shifts to inform conservation efforts. Here we showed that extinctions due to climate-induced range shifts are more likely in populations structured by steeper environmental gradients or starker range edges. For our approach, we developed an individual-based model to explore the eco-evolutionary dynamics of climate-induced range shifts in sexually reproducing, diploid populations with both dispersal and niche traits defined by multiple loci. In contrast, previous models have focused on a subset of these factors: ecological dynamics (e.g.~\citep{brooker2007modelling}), evolution in a single trait only (e.g.~\citep{atkins2010local, henry2013eco}), and relatively simple genetic scenarios (e.g. single-locus haploid genetics in asexual populations~\citep{boeye2013more, hargreaves2015fitness}).  Previous studies on eco-evolutionary dynamics of populations responding to climate change have generally predicted that evolution of increased dispersal might provide relief for populations struggling to keep pace with changing environmental conditions~\citep{boeye2013more, henry2013eco, hargreaves2015fitness}. Here, we tested the generality of these predictions in populations with varying types of spatial structure induced by the underlying environmental conditions.

While all surviving populations evolved increased dispersal phenotypes during range shifts, the environmental gradient and edge starkness may have hampered the ability of populations to achieve such increases in dispersal under certain scenarios. For example, scenarios characterized by stark edges or steep environmental gradients led to edge populations composed of lower dispersal phenotypes at equilibrium due to the increased cost of dispersal in these scenarios. In the presence of a steep environmental gradient, selection favors lower dispersal phenotypes due to the fitness cost of dispersing to a location with a drastically different environmental optimum~\citep{kirkpatrick1997evolution}. Similarly, stark range edges can select for lower dispersal due to the increased risk of dispersing into unfavorable habitat~\citep{shaw2014population, shaw2019evolution}. Previous research has documented a similar reduction in dispersal phenotypes due to an explicit mortality cost of dispersal~\citep{kubisch2013predicting}, whereas the costs to dispersal in our model result from the environmental gradient and range edge starkness. By starting from lower dispersal phenotypes, populations structured by steep environmental gradients or stark range edges needed to achieve greater evolutionary increases in dispersal, since populations in all scenarios evolved to similar levels of dispersal in response to climate change (Fig. 3). Presumably, the increased magnitude of the required evolutionary change in dispersal might have contributed to the increased extinction risk observed in such scenarios.

Additionally, both the environmental gradient and the edge starkness impacted the spatial distribution of genetic variance in dispersal at equilibrium, thus altering the evolutionary potential of populations in different scenarios. Equilibrium genetic variance in dispersal decreased with an increasingly steep environmental gradient. Coupled with the reduced dispersal phenotypes characteristic of these scenarios at equilibrium, this reduction in genetic diversity could have amplified the difficulty of populations evolving the increased dispersal necessary to track climate change. While the effect of edge type on equilibrium patterns of genetic variance in dispersal was less straightforward, resulting in qualitative differences among different levels of edge starkness, it provides a clue to the lower extinction risk faced by scenarios defined by moderate versus gradual edges (Fig. 2). Scenarios defined by moderate edges consistently had higher levels of genetic variance in dispersal compared to scenarios with gradual edges, suggesting the evolution of increased dispersal during range shifts may have been easier in such scenarios.
